---
title: "Homework Assignment 2"
author: "Ixel"
format: pdf
editor: visual
---

## Exploring patterns of environmental justice

#### Set Up

```{r}
# Librarys
rm(list = ls()) # whats this for?
library(sf) # for vector data 
library(stars) # for raster data 
library(tmap) # for static and interactive maps
library(here)
library(spData) # preloaded spatial data what does that mean?
library(tidyverse)
```

```{r}
# spatial data objest to plot

# raster of global elevations
ejscreen <- read_sf(here("data/ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb",))

# country boundaries
... <- read_sf(here("data",))

# major cities
...  <- read_sf(here("data",))
```


#### Part 1: Legacy of redlining in current environmental (in)justice
 
Exploring historical redlining in Los Angeles and its legacy on present-day environmental justice.
 
map of historical redlining neighborhoods, including:
- neighborhoods colored by HOLC grade
- an appropriate base map
- a clear title and legend
 
```{r}

```

Part 2: Legacy of redlining in biodiversity observations